---
title: "Ignitions_FWI_11_2020"
author: "Zjrobbin"
date: "11/19/2020"
output: github_document 
---


```{r}
library(RColorBrewer)
library(pscl)
library(plyr)
```

#### FWI~ Ignition relationship 
The fire weather index and igntion relationship was updated in Nov 2020
I created to script to better explain the new choices available within the
code, code is still based on AK 2017 work -ZR

Here is an example of the input file 
https://github.com/LANDIS-II-Foundation/Extension-SCRPPLE/blob/master/Testing/Core7-SCRAPPLE2.6/scrapple.txt

The user currently has two choices in shape in fititng the data the ignition algorithim. 

* Poisson
* ZeroInflated Poisson 

This choice is entered on line 29
If ZeroInflatedPoisson is chosen an additional four parameters are needed.


* LightningIgnitionsBinomialB0
* LightningIgnitionsBinomialB1  
* AccidentalIgnitionsBinomialB0 
* AccidentalIgnitionsBinomialB1 


## Example

For this example we will be using data from Karen Shorts Fire Database from a dataset used for the Southern Applachain Landsacpe 

** Short, Karen C. 2017. Spatial wildfire occurrence data for the United States, 1992-2015 [FPA_FOD_20170508]. 4th Edition. Fort Collins, CO: Forest Service Research Data Archive. https://doi.org/10.2737/RDS-2013-0009.4
https://www.fs.usda.gov/rds/archive/Product/RDS-2013-0009.4/ ** 

Abstract:This data publication contains a spatial database of wildfires that occurred in the United States from 1992 to 2015. It is the third update of a publication originally generated to support the national Fire Program Analysis (FPA) system. The wildfire records were acquired from the reporting systems of federal, state, and local fire organizations. The following core data elements were required for records to be included in this data publication: discovery date, final fire size, and a point location at least as precise as Public Land Survey System (PLSS) section (1-square mile grid). The data were transformed to conform, when possible, to the data standards of the National Wildfire Coordinating Group (NWCG). Basic error-checking was performed and redundant records were identified and removed, to the degree possible. The resulting product, referred to as the Fire Program Analysis fire-occurrence database (FPA FOD), includes 1.88 million geo-referenced wildfire records, representing a total of 140 million acres burned during the 24-year period.

[metadata](https://www.fs.usda.gov/rds/archive/products/RDS-2013-0009.4/_metadata_RDS-2013-0009.4.html)


These data were spatially located then divided into lightning ignitions and accidental-human ignited values. 



```{r setup, include=FALSE}
#w_dir <- "C:/Users/zacha/Desktop/Truncated_Scrapple/Inputs/"
l_fire_dat <- read.csv("FiresInAppsLightning2.csv")
head(l_fire_dat)
l_fire_days <- as.data.frame(cbind(l_fire_dat$FIRE_YEAR, l_fire_dat$DISCOVER_1)) #Extracting year and julian day
colnames(l_fire_days) <- c("YEAR", "J_DAY")
l_fire_days_sort <- l_fire_days[order(l_fire_days[,1]),] #sorting b
```


Here we can see the number of lightning ignited fires per year


```{r}
red<-RColorBrewer::brewer.pal(9,'YlOrRd')
#plot no of fires/yr
l_fire_days_sort_count<- cbind(l_fire_days_sort, rep(1, nrow(l_fire_days_sort)))
l_fires_count <- cbind.data.frame(unique(l_fire_days_sort_count$YEAR) , 
                                 tapply(l_fire_days_sort_count$`rep(1, nrow(l_fire_days_sort))`,l_fire_days_sort_count$YEAR, sum))
colnames(l_fires_count) <- c("YEAR", "COUNT")

barplot(l_fires_count$COUNT, main ="No of ign/yr Lightning Observed",col=red,names.arg=l_fires_count$YEAR)

```


and the human Ignited 

```{r}
h_fire_dat <- read.csv("FiresInAppsHuman2.csv")
#head(h.fire.dat)
h_fire_days <- as.data.frame(cbind(h_fire_dat$FIRE_YEAR, h_fire_dat$DISCOVER_1)) #Extracting year and julian day
colnames(h_fire_days) <- c("YEAR", "J_DAY")
h_fire_days_sort <- h_fire_days[order(h_fire_days[,1]),] #sorting by year
#plot no of fires/yr
h_fire_days_sort_count<- cbind(h_fire_days_sort, rep(1, nrow(h_fire_days_sort)))
h_fires_count <- cbind.data.frame(unique(h_fire_days_sort_count$YEAR) , 
                                  tapply(h_fire_days_sort_count$`rep(1, nrow(h_fire_days_sort))`,h_fire_days_sort_count$YEAR, sum))
colnames(h_fires_count) <- c("YEAR", "COUNT")
barplot(h_fires_count$COUNT, main ="No of ign/yr Human Accidental Observed",col=red,names.arg=h_fires_count$YEAR)

```


Here we Mmtch them up with Fire weather index(FWI) data generated from the historic record of weather fed through LANDIS-II Fire weather to get the models calculation of FWI. See the Climate library documentation for assistance getting FWI. 


```{r}

ign_types <- c("Lightning", "HumanAccidental")

fire_days_list <- list(l_fire_days_sort, h_fire_days_sort) ##organizing all ignition types into a list


##Import daily historic FWI data
FWI_dat <- read.csv("Climate-future-input-log.csv")

FWI_dat<-with(FWI_dat,aggregate(FWI,by=list(Year=Year,Timestep=Timestep),FUN=mean))



FWI_dat <- FWI_dat[,c(1,2,3)]
colnames(FWI_dat)[3]<-"FWI"
FWI_dat$ID <- paste(FWI_dat$Year, "_", FWI_dat$Timestep, sep="") #creating date identifier out of date and julian day
igns_list <- list()

for (i in 1:length(ign_types[1:2])){#THIS DOESN'T INCLUDE RX BURNS BUT THATS CAUSE WE ARE PROVIDING THOSE TO SCRAPPLE DIRECT
  ign_type_select <- fire_days_list[[i]] ##selecting each ignition type individually
  fire_days_count <- ddply(ign_type_select, .(ign_type_select$YEAR, ign_type_select$J_DAY), nrow) #finds duplicate rows in fire data
  colnames(fire_days_count) <- c("YEAR", "JD", "No_FIRES") #Renaming columns for consistency
  fire_days_count$ID <- paste(fire_days_count$YEAR, "_", fire_days_count$JD, sep="") #creating date identifier out of date and julian day

  ##Merging dataframes by year and julian day
  fire_days_short <- subset(fire_days_count, fire_days_count$YEAR < 2014) ##restricting fire records to climate file years
  FWI_short <- subset(FWI_dat, FWI_dat$Year > 1991) #restricting climate data to fire history records
  merge_col <- FWI_short$ID
  FWI_fire_merge <- join(FWI_short, fire_days_short, type="left") ##Merging based on unique date id
  FWI_fire_number <- FWI_fire_merge[,c(1,3,7)] #pulling out FWI and number of fires
  FWI_fire_number[is.na(FWI_fire_number)] <- 0 #converting NAs to 0, meaning 0 fires
  plot(FWI_fire_number[,2], FWI_fire_number[,3], main =ign_types[i], xlab = "dailyFWI", ylab = "noFires") #plotting FWI against no of fires just to look at pattern
  igns_list[[i]] <- FWI_fire_number 
}
```


```{r}
#Lighting
print("Lighting")
summary(igns_list[[1]])
#Humnan
print("Human")
summary(igns_list[[2]])

```

Now we individual fit the two models: One for the Lightning and one for the Human Accidetnal. 

This is where we can decide whether our model will be Poisson or Zero Inflated Poisson:

You can read more about Poisson and Zero Inflate Poisson here. 
https://math.usu.edu/jrstevens/biostat/projects2013/rep_ZIP.pdf 

In the simpliest terms we are interested in whether the number of zeros causes the fixed varaince in the poisson model to make the model overdispersed. 
https://en.wikipedia.org/wiki/Overdispersion
https://stats.stackexchange.com/questions/66586/is-there-a-test-to-determine-whether-glm-overdispersion-is-significant

Here is both models fit to a Poisson 


```{r}
Lightning<-as.data.frame(igns_list[[1]])
POIS_mod_L <- glm(No_FIRES~FWI,data=Lightning, family="poisson")
summary(POIS_mod_L)
```

```{r}
Accidental<-as.data.frame(igns_list[[2]])
POIS_mod_A <- glm(No_FIRES~FWI,data=Accidental, family="poisson")
summary(POIS_mod_A)
```


Both of these parameters have two parameters so if you chose this model your parameter file would read   

(lines 24-29)


> LightningIgnitionsB0   -4.268433  
> LightningIgnitionsB1    0.135562   
> AccidentalIgnitionsB0   1.167798   
> AccidentalIgnitionsB1  -0.026771    

> IgnitionDistribution Poisson << Poisson or ZeroInflatedPoisson


and you would include nothing in the example until line 37 (Maximum Fine Fuels)


If you were unhappy with the fit of that model you can try a Zero Inflated Poisson

### Zero Inflated Poisson 




```{r}
Lightning<-as.data.frame(igns_list[[1]])

###Lightning
zeroinf_mod_L <- zeroinfl(No_FIRES~FWI,data=Lightning, dist="poisson")
summary(zeroinf_mod_L)
#Tst<-predict(zeroinf_mod,type="zero")
```



```{r}

Accidental<-as.data.frame(igns_list[[2]])
#Lightning$No_FIRES
###Lightning
zeroinf_mod_A<- zeroinfl(No_FIRES~FWI,data=Accidental, dist="poisson")
summary(zeroinf_mod_A)
```

These models each have four parameters so your input would look as follows   
From line (24-35)



> LightningIgnitionsB0   -0.63779  
> LightningIgnitionsB1   0.01623  
> AccidentalIgnitionsB0  1.920144   
> AccidentalIgnitionsB1  -0.021641  

> IgnitionDistribution ZeroInflatedPoisson << Poisson or ZeroInflatedPoisson  


> LightningIgnitionsBinomialB0 3.83717  
> LightningIgnitionsBinomialB1  -0.14822  
> AccidentalIgnitionsBinomialB0 0.115584   
> AccidentalIgnitionsBinomialB1  0.009369   



How these models impact your ignition model will be data dependent. 


### Testing 

Here we can test how those formulas will play out within the model 


```{r,fig.width=10.0,fig.height=7.0}
FWI_dat<- read.csv("Climate-future-input-log.csv")
FWI_dat<-with(FWI_dat,aggregate(FWI,by=list(Year=Year,Timestep=Timestep),FUN=mean))

FWI_dat<-FWI_dat[FWI_dat$Year %in% seq(1992,2016),]
FWI_Test<-FWI_dat$x
Pb0<-(1.167798)
Pb1<-(-.026771)
###Possion model 
Acc_Ign_Test_Pois<-rpois(9125, lambda = exp((Pb0+(Pb1*FWI_Test))))

##ZIPS model 


binomb0=.115584
binomb1=.009369

b0=(1.920144)
b1=(-.021641)

alpha<-exp(binomb0+binomb1*FWI_Test)
zeroprob= alpha/(alpha+1)
Acc_Ign_Test<-ifelse((zeroprob<=runif(length(zeroprob),0,1)),
        round(rpois(9125, lambda = exp((b0+(b1*FWI_Test))))),
       0)

### Aggregate to the year 
Acc_Ign_df<-cbind(FWI_dat[,c(1,2)],Acc_Ign_Test)
Acc_Years<-aggregate(list(Ignitions=Acc_Ign_df$Acc_Ign_Test),by=list(Year=Acc_Ign_df$Year),FUN=sum)

PAcc_Ign_df<-cbind(FWI_dat[,c(1,2)],Acc_Ign_Test_Pois)
PAcc_Years<-aggregate(list(Ignitions=PAcc_Ign_df$Acc_Ign_Test_Pois),by=list(Year=PAcc_Ign_df$Year),FUN=sum)

plot(PAcc_Years$Year,PAcc_Years$Ignitions,col="darkgreen",ylim=c(400,1500),pch=16,cex=1.3,xlab="Accidental Ignition",ylab="Year")
points(Acc_Years$Year,Acc_Years$Ignitions,col="purple",pch=16,cex=1.3)
points(h_fires_count$YEAR,h_fires_count$COUNT,col="orange",pch=16,cex=1.3)

legend(1995,600,legend=c("Accidental Possion","Accidental ZIPS","Karen Short"),pch=c(16,16,16),col=c("darkgreen","purple","orange"))
```


We can see there is not a huge differnce but perhaps the ZIPS model has a larger variance annually. However there is perhaps more variation in the realworld data than is captured in the model. 


Here we can compare to the results we get from the model 


```{r,fig.width=10.0,fig.height=7.0}
ZIP_Ignitions<-read.csv("scrapple-ignitions-log.csv")
Pois<-read.csv("scrapple-ignitions-logPois_2.csv")

ZIP<-aggregate(x=list(Count=ZIP_Ignitions$ActualNumberIgnitions,OtherCount=ZIP_Ignitions$AttemptedNumberIgnitions),
              by=list(Year=ZIP_Ignitions$SimulationYear,Type=ZIP_Ignitions$IgnitionType),
              FUN=sum)
P_inf<-aggregate(x=list(Count=Pois$ActualNumberIgnitions,OtherCount=Pois$AttemptedNumberIgnitions),
              by=list(Year=Pois$SimulationYear,Type=Pois$IgnitionType),
              FUN=sum)

#OIL<-OI[OI$Type==" Accidental",]

ZIPA<-ZIP[ZIP$Type==" Accidental",]
P_infA<-P_inf[P_inf$Type==" Accidental",]
#unique(Merged$IgnitionType)
plot(PAcc_Years$Year,ZIPA$Count[1:25],ylim=c(400,1500),pch=16,col="red",main="LANDIS-Accidental",ylab="Accidental Ignition",xlab="Year")
points(PAcc_Years$Year,P_infA$Count[1:25],col="blue",pch=16)
points(PAcc_Years$Year,PAcc_Years$Ignitions,col="darkgreen",pch=16,cex=1.3,xlab="Accidental Ignition",ylab="Year")
points(Acc_Years$Year,Acc_Years$Ignitions,col="purple",pch=16,cex=1.3)
points(h_fires_count$YEAR,h_fires_count$COUNT,col="orange",pch=16,cex=1.3)

legend(1995,700,legend=c("Accidental Possion","Accidental ZIPS","Karen Short","LANDIS POISSON","LANDIS ZIPS"),pch=c(16,16,16,16,16),col=c("darkgreen","purple","orange",'blue','red'))
```

We see agian the ZIPs model has more variance, more akin to the the data. 

Here is the ligthning ignitoin


```{r,fig.width=10.0,fig.height=7.0}
FWI_dat<- read.csv("Climate-future-input-log.csv")
FWI_dat<-with(FWI_dat,aggregate(FWI,by=list(Year=Year,Timestep=Timestep),FUN=mean))

FWI_dat<-FWI_dat[FWI_dat$Year %in% seq(1992,2016),]
FWI_Test<-FWI_dat$x
Pb0<-(-4.268433)
Pb1<-(.135562)
###Possion model 
Lig_Ign_Test_Pois<-rpois(9125, lambda = exp((Pb0+(Pb1*FWI_Test))))

##ZIPS model 


binomb0=3.83717
binomb1=-.14822

b0=(-.63779)
b1=(.01623)

alpha<-exp(binomb0+binomb1*FWI_Test)
zeroprob= alpha/(alpha+1)
Lig_Ign_Test<-ifelse((zeroprob<=runif(length(zeroprob),0,1)),
        round(rpois(9125, lambda = exp((b0+(b1*FWI_Test))))),
       0)

### Aggregate to the year 
Lig_Ign_df<-cbind(FWI_dat[,c(1,2)],Lig_Ign_Test)
Lig_Years<-aggregate(list(Ignitions=Lig_Ign_df$Lig_Ign_Test),by=list(Year=Lig_Ign_df$Year),FUN=sum)

PLig_Ign_df<-cbind(FWI_dat[,c(1,2)],Lig_Ign_Test_Pois)
PLig_Years<-aggregate(list(Ignitions=PLig_Ign_df$Lig_Ign_Test_Pois),by=list(Year=PLig_Ign_df$Year),FUN=sum)

plot(PLig_Years$Year,PLig_Years$Ignitions,col="darkgreen",ylim=c(0,100),pch=16,cex=1.3,ylab="Lightning Ignition",xlab="Year")
points(Lig_Years$Year,Lig_Years$Ignitions,col="purple",pch=16,cex=1.3)
points(l_fires_count$YEAR,l_fires_count$COUNT,col="orange",pch=16,cex=1.3)

legend(1995,100,legend=c("Lightning Possion","Lightning ZIPS","Karen Short"),pch=c(16,16,16),col=c("darkgreen","purple","orange"))
```
 
and with the LANDIS Scrpple lightning outputs using the above parameters


```{r,fig.width=10.0,fig.height=7.0}
ZIPL<-ZIP[ZIP$Type==" Lightning",]
P_infL<-P_inf[P_inf$Type==" Lightning",]

plot(PAcc_Years$Year,ZIPL$Count[1:25],ylim=c(0,100),pch=16,col="red",main="LANDIS-Lightning",xlab="Lightning Ignition",ylab="Year")
points(PAcc_Years$Year,P_infL$Count[1:25],col="blue",pch=16)
points(PLig_Years$Year,PLig_Years$Ignitions,col="darkgreen",pch=16,cex=1.3,xlab="Accidental Ignition",ylab="Year")
points(Lig_Years$Year,Lig_Years$Ignitions,col="purple",pch=16,cex=1.3)
points(l_fires_count$YEAR,l_fires_count$COUNT,col="orange",pch=16,cex=1.3)

legend(1995,100,legend=c("Accidental Possion","Accidental ZIPS","Karen Short","LANDIS POISSON","LANDIS ZIPS"),pch=c(16,16,16,16,16),col=c("darkgreen","purple","orange",'blue','red'))
```

